Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
context about science added

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{amsthm}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}
\usepackage{subcaption}
\usepackage{siunitx} % For pretty-printing numeric values and SI units
                     % of measure. e.g., the tabular column type S is
                     % used to print nice-looking tables of numbers.
\sisetup{ % defaults
  group-separator={,},
  group-minimum-digits={3},
  output-decimal-marker={.},
  table-format = 6
}

@@

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in the middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical code, but
  one which is error-prone due to fine-grained indexing errors. We
  hypothesise that in practise, stencil computations tend to have a
  regular shape which is amenable to static analysis and simple
  specification. We describe a abstract spatial specification language for
  stencil computations that can be embedded as annotations in
  source code, along with an inference and checking procedure. We
  evaluate our language against a corpus of numerical Fortran code,
  and show that precise specifications can be given to roughly half of
  the identifiable stencil computations, and that the vast majority of
  stencil computations indeed have a simple, regular, static
  shape. This paper details our stencil specification language, its
  static semantics, inference, checking, and program synthesis
  procedures, our implementation, and our evaluation study. 
\end{abstract}

@@

In this work, we introduce a simple specification language for the
spatial properties of stencil computations. The
specifications abstract over the fine grained detail of stencil access
patterns. In practice, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{minted}{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
pictorially.

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce a specification language for
stencil computations that captures many common forms
of data access pattern, both spatial and temporal (Section~\ref{sec:lang});
\item we detail inference and checking
algorithms for stencil specifications (Section~\ref{sec:analysis});
\item we evaluate our approach, providing an 
implementation as an extension of CamFort,
an open-source program analysis and refactoring tool for Fortran
corpus. We use this implementation to evaluate our approach against
a corpus of scientific applications, totalling roughly one million
lines of code, and we study some verification case studies.
\dnote{insert results here}
\end{itemize}
%

\begin{figure}[t]
\begin{minted}[firstnumber=20]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+ &
  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))- &
  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
  /(4.0*delx)

@@

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{minted}
(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. 
\vspace{0.5em}
\hrule
%
\begin{minted}{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred stencil specification from CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.3em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

Our specification system is based on the observation
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
``\emph{five-point stencil}'' on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} where the inference of
such regular stencil patterns on a corpus of numerical programs (both
small and large). We found that indeed \dnote{..}.

We outline the specification language here. Section~\ref{sec:syntax}
outlines the syntax. Section~\ref{sec:semantics} defines its semantics
via a simple set interpretation over indices. Section~\ref{sec:eqs} provides an
equational theory for specifications via relation ($\equiv$) and a
theory of approximation via relation ($<:$). These equations are then
proven sound with respect to the set semantics of Section~\ref{sec:semantics}.

@@

\renewcommand*{\arraystretch}{0.8}
\paragraph{Target language syntax} Throughout, for the source language
syntax, $v$ ranges over program variables, $s$ over statements, and
$e$ over expressions (which may be impure).  By ``variable'', we mean
in the imperative sense (\ie{}, named binders to memory cells) instead
of mathematical variables. We make clear a number of standard notions
used throughout.

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop\footnote{Or equivalent for the target language, \eg{},
    \texttt{do} in Fortran for our implementation}, incremented by $1$ per
iteration. The variable is marked as a base induction variable
only within the scope of the loop, and such variables will be ranged
over by $i$, $j$, $k$. Since we do not use non-base induction
variables (which are traditionally affine transformations of induction
variables) we often simply say \emph{induction variable}.
\end{definition}

\begin{definition}[Array subscript]
  An \emph{array subscript} is an expression that indexes an array
  (for the purpose or reading or writing an element), which we denote
  by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
  for a syntactic list of indexing expressions. A \emph{relative
    index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
  in terms of some base induction variable.
\end{definition}

\begin{definition}[Neighbourhood index and neighbourhood subscripts]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is a
  constant of integer type. That is, a neighbourhood index is a
  relative index defined as a constant translation of a base induction
  variable. The relation $\equiv$ identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. An array subscript $v(\bar{e})$ where every
 $e \in \bar{e}$ is a neighbourhood index is called a
 \emph{neighbourhood subscript}.
\label{def:neighbour}
\end{definition}

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{amsthm}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}
\usepackage{subcaption}
%\usepackage{subfigure}
\usepackage{siunitx} % For pretty-printing numeric values and SI units
                     % of measure. e.g., the tabular column type S is
                     % used to print nice-looking tables of numbers.
\sisetup{ % defaults
  group-separator={,},
  group-minimum-digits={3},
  output-decimal-marker={.},
  table-format = 6
}

@@

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in the middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical
  code, but one which is error-prone due to fine-grained indexing
  errors. We hypothesise that in practise, stencil computations tend
  to have a regular shape which is amenable to static analysis and
  simple specification. We describe a abstract spatial specification
  language for stencil computations that can be embedded as
  annotations in source code, along with inference, checking, and
  specification and program synthesis procedures. We evaluate our
  language against a corpus of numerical Fortran code, and show that
  precise specifications can be given to roughly half of the
  identifiable stencil computations, and that the vast majority of
  stencil computations indeed have a simple, regular, static
  shape. This paper details our stencil specification language, its
  static semantics, inference, checking, and program synthesis
  procedures, our implementation, and our evaluation study.
\end{abstract}

@@

In practice, most stencil computations have a regular shape that can
be described simply and abstractly with a small set of coarse-grained
spatial descriptions. We introduce a simple specification language of
such descriptions, abstracting over the fine-grained detail of stencil
access patterns. For our first Laplace example, our
inference procedure provides the following specification to line 3:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This explains that \texttt{a} is read from with a
symmetrical access pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) illustrates
the specifications pictorially.

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce our specification language for
stencil computations that captures many common forms
of data access pattern (Section~\ref{sec:lang})
along with a denotational model (Section~\ref{sec:semantics});

\item we detail inference, checking, specification synthesis, 
and program synthesis algorithms for our language
(Section~\ref{sec:analysis}), which relies on the semantic model;

\item we evaluate our approach, providing an 
implementation as an extension of CamFort,
an open-source program analysis and refactoring tool for Fortran
corpus. We use this implementation to evaluate our approach against
a corpus of scientific applications, totalling roughly one million
lines of code, and we study some verification case studies.
\dnote{insert results here}
\end{itemize}
%

\begin{figure}[t]
\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+ &
  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))- &
  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
  /(4.0*delx)

@@

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{minted}
(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. 
\vspace{0.5em}
\hrule
%
\begin{minted}[linenos=false,xleftmargin=1em]{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred stencil specification from CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}[scale=0.85]
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.3em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language}
\label{ref:navier-stokes-fragment}
\end{figure}

\paragraph{Context}

In recent years, there has been an increased awareness 
of the need for formal verification techniques in 
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}. 
Our specification language is an approach in this direction, proving 
a system of lighweight specification for numerical code.
The aim is to not only provide a way of exposing errors, but also
for specifying program behaviour to future-proof the code and ease
its maintenance. In order to increase the likelihood of impact, we aim
to integrate with existing practises smoothly, hence our
specifications are integrated as comments and our tool can synthesis
specifications for the user.

Much of this paper concerns the technical aspects of the language,
its semantics, its inference, checking, and synthesis procedures.
The evaluation focuses on its application to real scientific code base
used actively in research.

@@

Our specification language is based on the observation
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
``\emph{five-point stencil}'' on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} where the inference of
such regular stencil patterns on a corpus of numerical programs (both
small and large). We found that indeed \dnote{..}.

Section~\ref{sec:syntax} outlines the syntax of specifications. 
 Section~\ref{sec:eqs} provides an equational
theory for specifications via relation ($\equiv$) and a notion of
sub-specification via relation ($<:$).  Section~\ref{sec:semantics}
gives a semantic model via a set-based interpretation, for which the
equality and sub-specification relation are proven sound.

@@

\renewcommand*{\arraystretch}{0.8}
\paragraph{Target language syntax} Throughout, for the source language
syntax, $v$ ranges over program variables, $s$ over statements, and
$e$ over expressions (which may be impure).  By ``variable'', we mean
in the imperative sense (\ie{}, named binders to memory cells) instead
of mathematical variables.

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop,\footnote{Or equivalent for the target language, \eg{},
    \texttt{do} in Fortran for our implementation} incremented by $1$ per
iteration. The variable is marked as a base induction variable
only within the scope of the loop. Induction variables are ranged
over by $i$, $j$, $k$ throughout. Since we do not use non-base
(derived) induction variables we often simply say \emph{induction variable}.
\end{definition}

\begin{definition}[Array subscript and index]
  An \emph{array subscript} is an expression that indexes an array
  (for the purpose of reading or writing an element), which we denote
  by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
  for a syntactic list of indexing expressions. A \emph{relative
    index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
  in terms of base induction variables.
\end{definition}

\begin{definition}[Neighbourhood index and subscript]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
  integer constant. That is, a neighbourhood index is a
  relative index defined as a constant translation of a base induction
  variable. The relation $\equiv$ identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. An array subscript $v(\bar{e})$ where every
 $e \in \bar{e}$ is a neighbourhood index is called a
 \emph{neighbourhood subscript}.
\label{def:neighbour}
\end{definition}
